---
title: "Assignment 3, Task 1: Principal Components Analysis"
author: "I. Requena"
date: "2023-02-28"
output: 
  html_document:
    code_folding: hide
---

### Introduction:

The purpose of Task 1 of Assignment 3 is to conduct Principal Components Analysis (PCA) using [world environmental data](https://www.kaggle.com/datasets/zanderventer/environmental-variables-for-world-countries). The data was constructed through publicly available remote sensing datasets provided by [Google Earth Engine](https://earthengine.google.com/). With this information, we want to determine to what degree do elevation, temperature, rainfall, and cloud affect biomass (surface cropland and tree canopy surface). Furthermore, we would explore how these variables related.

```{r setup, echo=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, results=FALSE)

library(tidyverse)
library(here)


# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(patchwork)

# For Part 2 (ggplot customization)
library(readxl)
library(gghighlight) # For useful highlighting
library(lubridate) # For easier dates & times
library(plotly) # Interactive graphs
```

### PCA biplot

The lenght of a line indicates variance (longer length=more variance)
The angle between lines indicates correlation

Crop cover and elevation are positively correlated. 

```{r}
# 1. Read & Wrangle data

df_global <- read_csv(here('data/world_env_vars.csv')) %>%
  drop_na() %>% 
  select(elevation, cropland_cover, tree_canopy_cover, rain_mean_annual, temp_mean_annual, cloudiness)

global_pca <- df_global %>%
  dplyr::select(where(is.numeric)) %>%  # selects when its numeric across all different columns
  scale() %>% #takes each column with numeric values (life expectancy and rescaling by the average)
  prcomp()

#2. Create biplot automatically with ggfortify

autoplot(global_pca,
     	data = df_global, 
     	loadings = TRUE,
     	loadings.label = TRUE,
     	loadings.colour = "blue",
     	loadings.label.colour = "darkgrey",
     	loadings.label.vjust = -0.5)+
  theme_minimal()
```

### Screeplot

Useful for visualizing PC contributions

```{r}
# Variance explained by each PC
screeplot(global_pca, type = "lines")
screeplot(global_pca, type = "barplot")
```
### Summary

