---
title: "Assignment 3, Task 1"
author: "I. Requena"
date: "2023-02-28"
output: 
  html_document:
    code_folding: hide
---

### Introduction:

#1.	A useful descriptive introductory summary (3 - 4 sentences) that helps the audience understand the data (include a citation as necessary) and what you’ll be exploring

This is a selected set of variables:

In Task 1 of Assignment 3, we want to know if the morphological traits of palmetto palms can predict the species. To answer this question, a model was built which analyses morphological variables (such as height, canopoy length, canopy width, and leaf count) given the type of palm species (Serenoa repens or Sabal etonia). A second model was built with all variables but excluded canopy length. The data used in this task, obtained from the [EDI Data Portal](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.317.1), is part of a study conducted in south-central Florida from 1981 to 2017 in 5-year intervals (Abrahamson, 2019).

**Data source:** Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative.

```{r setup, echo=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, results=FALSE)

library(tidyverse)
library(here)


# For Part 1 (PCA)
library(ggfortify) # For PCA biplot
library(patchwork)

# For Part 2 (ggplot customization)
library(readxl)
library(gghighlight) # For useful highlighting
library(lubridate) # For easier dates & times
library(plotly) # Interactive graphs
```

### Read & Wrangle data
```{r}
df_global <- read_csv(here('data/world_env_vars.csv')) %>%
  drop_na() 

glimpse(df_global)

global_pca <- df_global %>%
  dplyr::select(where(is.numeric)) %>%  # selects when its numeric across all different columns
  scale() %>% #takes each column with numeric values (life expectancy and rescaling by the average)
  prcomp()
```

# Create a BiPlot automatically!
```{r}
autoplot(global_pca,
     	data = df_global,
     	loadings = TRUE,
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5)+
  scale_color_manual(values = c('red', 'orange', 'yellowgreen', 'darkgreen'))+
  theme_minimal()
```














